---
title: "Superstore_project"
output: html_document
date: "2024-08-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install necessary packages (if not already installed)
if (!requireNamespace("lubridate", quietly = TRUE)) install.packages("lubridate")
if (!requireNamespace("gridExtra", quietly = TRUE)) install.packages("gridExtra")
if (!requireNamespace("wordcloud2", quietly = TRUE)) install.packages("wordcloud2")
if (!requireNamespace("wordcloud", quietly = TRUE)) install.packages("wordcloud")
if (!requireNamespace("tm", quietly = TRUE)) install.packages("tm")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("forcats", quietly = TRUE)) install.packages("forcats")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")

# Load necessary libraries
library(wordcloud2)
library(wordcloud)
library(tm)
library(dplyr)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(forcats)
library(scales)

# Load the data
data <- read.csv('/Users/edoardochiari/Desktop/Superstore_data.csv')
```

## Data info and cleaning 

```{r}
# Display the structure of the dataset
str(data)
# Get summary statistics for the dataset
summary(data)
```

### Extra columns removal
```{r}
# Drop columns Row ID, Order ID, Ship Date, Customer ID, Customer Name,Product ID,Country,Postal Code
data <- data[, !names(data) %in% c("Row.ID", "Order.ID","Ship.Date","Customer.ID", "Customer.Name", "Product.ID", "Country", "Postal Code")]
```

### Removal for duplicates and null variable
```{r}
# Count missing values in each column
missing_counts <- colSums(is.na(data))
print(missing_counts)

# Identify duplicated rows and display them
duplicates <- data[duplicated(data), ]
print(duplicates)

# Remove subsequent duplicates while keeping the first occurrence
data <- data[!duplicated(data), ]

# Verify the dataset after removal
summary(data)
```
We can see that now the lenght of each variable has lowered by one, meaning that the first istance has been kept while the subsequent duplicate has been removed.

### Trend of Sales, Profits and Discounts
```{r}
# Convert 'Order Date' to Date format
data$Order.Date <- mdy(data$Order.Date)
str(data$Order.Date)

# Aggregate sales data by month
monthly_sales <- data %>%
  group_by(Month = floor_date(`Order.Date`, "month")) %>%
  summarise(Sales = sum(Sales))

monthly_discount <- data %>%
  group_by(Month = floor_date(`Order.Date`, "month")) %>%
  summarise(Discount = sum(Discount))

monthly_profit <- data %>%
  group_by(Month = floor_date(`Order.Date`, "month")) %>%
  summarise(Profit = sum(Profit))

# Calculate average values
avg_sales <- mean(monthly_sales$Sales)
avg_discount <- mean(monthly_discount$Discount)
avg_profit <- mean(monthly_profit$Profit)

# Aggregate sales data by year
yearly_sales <- data %>%
  group_by(Year = floor_date(`Order.Date`, "year")) %>%
  summarise(Sales = sum(Sales))

yearly_discount <- data %>%
  group_by(Year = floor_date(`Order.Date`, "year")) %>%
  summarise(Discount = sum(Discount))

yearly_profit <- data %>%
  group_by(Year = floor_date(`Order.Date`, "year")) %>%
  summarise(Profit = sum(Profit))

# Create the line plot for monthly sales
plot_monthly_sales <- ggplot(monthly_sales, aes(x = Month, y = Sales)) +
  geom_line(color = "#00F5D4") +  # Line color
  geom_point(color = "#00F5D4") +  # Marker color
  geom_hline(yintercept = avg_sales, color = "black", linetype = "dashed") +  # Add average sales line
  labs(title = "Monthly Sales Trend", x = "Order Date", y = "Sales") +  # Add titles and labels
  theme_minimal() +  # Use a minimal theme
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5))  # Add grid lines

# Create the line plot for monthly discount 
plot_monthly_discount <- ggplot(monthly_discount, aes(x = Month, y = Discount)) +
  geom_line(color = "#FEE440") +  
  geom_point(color = "#FEE440") +  
  geom_hline(yintercept = avg_discount, color = "black", linetype = "dashed") +  
  labs(title = "Monthly Discount Trend", x = "Order Date", y = "Discount") +  
  theme_minimal() + 
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5)) 

# Create the line plot for monthly profit
plot_monthly_profit <- ggplot(monthly_profit, aes(x = Month, y = Profit)) +
  geom_line(color = "#00BBF9") +  
  geom_point(color = "#00BBF9") +  
  geom_hline(yintercept = avg_profit, color = "black", linetype = "dashed") +  
  labs(title = "Monthly Profit Trend", x = "Order Date", y = "Profit") +
  theme_minimal() +  
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5)) 

# Create the line plot for the yearly sales
plot_yearly_sales <- ggplot(yearly_sales, aes(x = Year, y = Sales)) +
  geom_line(color = "#00F5D4") +  # Line color
  geom_point(color = "#00F5D4") +  # Marker color
  labs(title = "Yearly Sales Trend", x = "Order Date", y = "Sales") +  # Add titles and labels
  theme_minimal() +  # Use a minimal theme
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5))  # Add grid lines

# Create the line plot for the yearly disocount
plot_yearly_discount <- ggplot(yearly_discount, aes(x = Year, y = Discount)) +
  geom_line(color = "#FEE440") +  
  geom_point(color = "#FEE440") +  
  labs(title = "Yearly Discount Trend", x = "Order Date", y = "Discount") +  
  theme_minimal() + 
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5)) 

# Create the line plot for the yearly profit
plot_yearly_profit <- ggplot(yearly_profit, aes(x = Year, y = Profit)) +
  geom_line(color = "#00BBF9") +  
  geom_point(color = "#00BBF9") +  
  labs(title = "Yearly Profit Trend", x = "Order Date", y = "Profit") +
  theme_minimal() +  
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        plot.title = element_text(hjust = 0.5)) 

```

```{r, echo=FALSE}
# Display the plot
print(plot_monthly_sales)
print(plot_monthly_discount)
print(plot_monthly_profit)
print(plot_yearly_sales)
print(plot_yearly_discount)
print(plot_yearly_profit)
```

From these graphs we could answer various business related questions.

1. Which month had the highest monthly Sales, Discount and Profit?

The highest values for the three categories seems to be achieved in November 2017
```{r}
# Filter the dataset for November 2017 and summarize Sales, Discount, and Profit
november_2017_summary <- data %>%
  filter(format(Order.Date, "%Y-%m") == "2017-11") %>%
  summarise(
    November_2017_Sales = sum(Sales, na.rm = TRUE),
    November_2017_Discount = sum(Discount, na.rm = TRUE),
    November_2017_Profit = sum(Profit, na.rm = TRUE)
  )

# Display the result
print(november_2017_summary)
```

2. Which month had the lowest monthly Sales, Discount and Profit?

  - Lowest monthly Sales: February 2014
  - Lowest monthly Discount: February 2016
  - Lowest monthly Profit: January 2015 (Loss)
```{r}
# Calculate Sales for February 2014
february_2014_sales <- data %>%
  filter(format(Order.Date, "%Y-%m") == "2014-02") %>%
  summarise(February_2014_Sales = sum(Sales, na.rm = TRUE))

# Calculate Discount for February 2016
february_2016_discount <- data %>%
  filter(format(Order.Date, "%Y-%m") == "2016-02") %>%
  summarise(February_2016_Discount = sum(Discount, na.rm = TRUE))

# Calculate Profit for January 2015
january_2015_profit <- data %>%
  filter(format(Order.Date, "%Y-%m") == "2015-01") %>%
  summarise(January_2015_Profit = sum(Profit, na.rm = TRUE))

# Combine all results into a single data frame
summary_results <- cbind(february_2014_sales, february_2016_discount, january_2015_profit)

# Display the combined result
print(summary_results)
```

3. Which year showed the most growth in Sales, Discount and Profit?

The year that showed the most growth it's the year 2017 

4. What is monthly average of sales, Discount And Profit in 2015 and 2017?
```{r, message=FALSE}
# Filter data for the years 2015 and 2017
data_2015_2017 <- data %>%
  filter(format(Order.Date, "%Y") %in% c("2015", "2017"))

# Group by year and month and then calculate monthly sums
monthly_summary <- data_2015_2017 %>%
  group_by(Year = format(Order.Date, "%Y"), Month = format(Order.Date, "%m")) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Discount = sum(Discount, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE)
  ) %>%
  ungroup()

# Calculate the monthly averages for each year
monthly_averages <- monthly_summary %>%
  group_by(Year) %>%
  summarise(
    Avg_Sales = mean(Total_Sales),
    Avg_Discount = mean(Total_Discount),
    Avg_Profit = mean(Total_Profit)
  )

# Display the monthly averages for 2015 and 2017
print(monthly_averages)
```

### Sales, Profits and Discounts by state
```{r}
# Summarize the data by State for each metric
state_summary <- data %>%
  group_by(State) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Total_Discount = sum(Discount, na.rm = TRUE)
  )

# Bar plot for Sales by State (Flipped Axes, Reversed Order, and Formatted Y-Axis)
barplot_sales <- ggplot(state_summary, aes(x = fct_rev(State), y = Total_Sales)) +
  geom_bar(stat = "identity", fill = "#00F5D4", color = "black") +
  labs(title = "Total Sales by State", x = "State", y = "Sales") +
  theme_minimal() +
  coord_flip() +  # Flip the axes
  scale_y_continuous(labels = comma) +  # Format y-axis labels
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5))   # Adjust text size

# Bar plot for Profit by State 
barplot_profit <- ggplot(state_summary, aes(x = fct_rev(State), y = Total_Profit)) +
  geom_bar(stat = "identity", fill = "#00BBF9", color = "black") +
  labs(title = "Total Profits by State", x = "State", y = "Profits") +
  theme_minimal() +
  coord_flip() +  
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 
 

# Bar plot for Discount by State 
barplot_discount <- ggplot(state_summary, aes(x = fct_rev(State), y = Total_Discount)) +
  geom_bar(stat = "identity", fill = "#FEE440", color = "black") +
  labs(title = "Total Discounts by State", x = "State", y = "Discounts") +
  theme_minimal() +
  coord_flip() +  
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 

```

```{r, echo=FALSE}
# Print the bar plots
print(barplot_sales)
print(barplot_profit)
print(barplot_discount)

```

#### Top 10 States by Sales, Profits and Discounts
```{r}
# Filter the top 10 states by Sales, Profit, and Discount
top_10_sales <- state_summary %>% arrange(desc(Total_Sales)) %>% slice(1:10)
top_10_profit <- state_summary %>% arrange(desc(Total_Profit)) %>% slice(1:10)
top_10_discount <- state_summary %>% arrange(desc(Total_Discount)) %>% slice(1:10)

# Bar plot for Top 10 States by Sales
barplot_top_sales <- ggplot(top_10_sales, aes(x = fct_reorder(State, Total_Sales), y = Total_Sales)) +
  geom_bar(stat = "identity", fill = "#00F5D4", color = "black") +
  labs(title = "Top 10 States by Sales", x = "State", y = "Sales") +
  theme_minimal() +
  coord_flip() +  # Flip the axes
  scale_y_continuous(labels = comma) +  # Format y-axis labels
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 
  # Adjust text size

# Bar plot for Top 10 States by Profit
barplot_top_profit <- ggplot(top_10_profit, aes(x = fct_reorder(State, Total_Profit), y = Total_Profit)) +
  geom_bar(stat = "identity", fill = "#00BBF9", color = "black") +
  labs(title = "Top 10 States by Profits", x = "State", y = "Profits") +
  theme_minimal() +
  coord_flip() +  
  scale_y_continuous(labels = comma) +  
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 
  

# Bar plot for Top 10 States by Discount
barplot_top_discount <- ggplot(top_10_discount, aes(x = fct_reorder(State, Total_Discount), y = Total_Discount)) +
  geom_bar(stat = "identity", fill = "#FEE440", color = "black") +
  labs(title = "Top 10 States by Discounts", x = "State", y = "Discounts") +
  theme_minimal() +
  coord_flip() +  
  scale_y_continuous(labels = comma) + 
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 

```

```{r, echo=FALSE}
# Print the bar plots
print(barplot_top_sales)
print(barplot_top_profit)
print(barplot_top_discount)
```

General Conclusions from the Graphs:

- High Sales vs. Profitability:
States with the highest total sales (like Texas and California) don’t necessarily have the highest profits. In fact, Texas, despite having the highest sales, is experiencing significant losses, likely due to the substantial discounts offered. This suggests that high sales alone don’t guarantee profitability; effective cost management and pricing strategies are crucial.

- Impact of Discounts:
There appears to be a correlation between high discounts and lower profitability. States that offer significant discounts (e.g., Texas and Tennessee) tend to have lower profits or even losses, indicating that excessive discounting might erode profit margins. Conversely, states with moderate discounts (e.g., California and New York) tend to show better profitability, suggesting a more balanced and effective use of discounts.

- Profit Outliers:
States like California and New York, which are among the highest in terms of profit, have found a sweet spot between offering discounts and maintaining profitability. They manage to generate significant sales without overly relying on discounts, resulting in substantial profits.

- Low Sales and Profitability:
States with low total sales generally show low profits as well, which is expected. However, if these states also have low discounts, it might suggest that their sales volumes are too low to cover fixed costs, or that they might need to reconsider their pricing or marketing strategies to improve sales.

- Sales and Profit Distribution:
The data suggests a wide disparity in sales and profitability across states. A few states dominate in terms of both sales and profit, while many others lag significantly behind. This could indicate market saturation in top-performing states or untapped potential in underperforming ones.


Strategic Insights:

- Balancing Discounts and Profit:
Excessive discounting does not always lead to better financial outcomes. States that carefully balance discounts while focusing on driving sales through other means (e.g., better marketing, customer loyalty programs) tend to perform better in terms of profitability.

- Focus on Underperforming States:
States with low sales and profits might benefit from targeted strategies to boost market presence and customer engagement. This could involve tailored promotions, localized marketing campaigns, or reassessing product offerings to better meet local demand.

- Leverage High-Performing States:
States that are already performing well in terms of both sales and profit could serve as models for strategies that could be applied to other regions. Understanding what drives success in these states could help in replicating that success elsewhere.

#### Top 10 Cities by Sales, Profits and Discounts
```{r}
# Summarize the data by Cities for each metric
cities_summary <- data %>%
  group_by(City) %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(Profit, na.rm = TRUE),
    Total_Discount = sum(Discount, na.rm = TRUE)
  )

# Filter the top 10 Cities by Sales, Profit, and Discount
top_10_cities_sales <- cities_summary %>% arrange(desc(Total_Sales)) %>% slice(1:10)
top_10_cities_profit <- cities_summary %>% arrange(desc(Total_Profit)) %>% slice(1:10)
top_10_cities_discount <- cities_summary %>% arrange(desc(Total_Discount)) %>% slice(1:10)

# Bar plot for Top 10 States by Sales
barplot_top_city_sales <- ggplot(top_10_cities_sales, aes(x = fct_reorder(City, Total_Sales), y = Total_Sales)) +
  geom_bar(stat = "identity", fill = "#00F5D4", color = "black") +
  labs(title = "Top 10 Cities by Sales", x = "City", y = "Sales") +
  theme_minimal() +
  coord_flip() +  # Flip the axes
  scale_y_continuous(labels = comma) +  # Format y-axis labels
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) # Adjust text size

# Bar plot for Top 10 States by Profit
barplot_top_city_profit <- ggplot(top_10_cities_profit, aes(x = fct_reorder(City, Total_Profit), y = Total_Profit)) +
  geom_bar(stat = "identity", fill = "#00BBF9", color = "black") +
  labs(title = "Top 10 Cities by Profits", x = "City", y = "Profits") +
  theme_minimal() +
  coord_flip() +  
  scale_y_continuous(labels = comma) +  
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 

# Bar plot for Top 10 States by Discount
barplot_top_city_discount <- ggplot(top_10_cities_discount, aes(x = fct_reorder(City, Total_Discount), y = Total_Discount)) +
  geom_bar(stat = "identity", fill = "#FEE440", color = "black") +
  labs(title = "Top 10 Cities by Discounts", x = "City", y = "Discounts") +
  theme_minimal() +
  coord_flip() +  
  scale_y_continuous(labels = comma) + 
  theme(axis.text.y = element_text(size = 8, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 

```

```{r, echo=FALSE}
# Print the bar plots
print(barplot_top_city_sales)
print(barplot_top_city_profit)
print(barplot_top_city_discount)
```

### Other visualizations

#### Count of Category and Sub-category
```{r}
#Count the occurrences of each category
category_counts <- data %>%
  count(Category) %>%
  arrange(desc(n))

#Count the occurrences of each sub-category
subcategory_counts <- data %>%
  count(Sub.Category) %>%
  arrange(desc(n))

#Create the category barplot
category_barplot <- ggplot(category_counts, aes(x = reorder(Category, -n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Create bar plot
  labs(title = "Category Counts", x = "Category", y = "Count") +  # Add titles and labels
  theme_minimal() +  # Use a minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) # Rotate x-axis labels for readability

#Create the sub-category barplot
subcategory_barplot <- ggplot(subcategory_counts, aes(x = reorder(Sub.Category, -n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Sub-category Counts", x = "Sub-category", y = "Count") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) 

```

```{r, echo=FALSE}
# Display the plot
grid.arrange(category_barplot, subcategory_barplot, ncol = 2)
```

#### Distribution of Shipments

```{r}
#Calculate the counts and percentages of Ship Mode
ship_mode_counts <- data %>%
  count(Ship.Mode) %>%
  mutate(Percentage = n / sum(n) * 100)

# Step 2: Create the pie chart
pie_chart <- ggplot(ship_mode_counts, aes(x = "", y = Percentage, fill = Ship.Mode)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Distribution of Ship Mode", fill = "Ship Mode") +
  theme_void() +  # Remove background, grids, and axis lines
  theme( plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")),
            position = position_stack(vjust = 0.5)) +  # Add percentage labels inside the pie slices
  scale_fill_manual(values = c("#007d7c", "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c"))  #Define custom colors
```

```{r, echo=FALSE}
# Display the pie chart
print(pie_chart)
```

#### Count and Distribution of Customer Segment
```{r}
# Calculate the counts of the Segment variable
segment_counts <- data %>%
  count(Segment)

# Create a color palette with the right number of colors
num_segments <- nrow(segment_counts)
colors <- scales::hue_pal()(num_segments)

# Create the pie chart for Segment
pie_chart_segment <- ggplot(segment_counts, aes(x = "", y = n, fill = Segment)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +
  theme_void() +  # Clean background and remove axis labels
  geom_text(aes(label = paste0(round(n / sum(n) * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +  # Add percentage labels
  scale_fill_manual(values = colors)  # Dynamically generated colors

# Create the bar chart for Segment
bar_chart_segment <- ggplot(segment_counts, aes(x = Segment, y = n, fill = Segment)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = n), vjust = -0.3, size = 4) +  # Add values on top of the bars
  labs(y = "Count", x = "Segment") +
  theme_minimal() +
  scale_fill_manual(values = colors) +  # Use the same dynamic color palette
  theme(legend.position = "none")  # Remove the legend

```

```{r, echo=FALSE}
# Combine the pie chart and bar chart side by side
grid.arrange(pie_chart_segment, bar_chart_segment, ncol = 2)
```

#### Words Cloud of Product names
```{r}
# Convert 'Product.Name' column to UTF-8 encoding
data$Product.Name <- iconv(data$Product.Name, from = "latin1", to = "UTF-8")

# Convert the Product.Name column to a vector of words
product_names <- tolower(data$Product.Name)  # Convert to lowercase
product_names <- removePunctuation(product_names)  # Remove punctuation
product_names <- removeNumbers(product_names)  # Remove numbers
product_names <- stripWhitespace(product_names)  # Remove extra whitespace

# Create a corpus and a term-document matrix
corpus <- Corpus(VectorSource(product_names))
tdm <- TermDocumentMatrix(corpus)
matrix <- as.matrix(tdm)
word_freqs <- sort(rowSums(matrix), decreasing=TRUE)
df_word_freqs <- data.frame(word = names(word_freqs), freq = word_freqs)
```

```{r, echo=FALSE}
# Generate the word cloud
wordcloud2(data = df_word_freqs, size = 1.5, color = "random-light", backgroundColor = "white")
```
